Outline flowCore paper

### 2008-04-29 paper/outline.tex
Deepayans suggestions and Application section
git-svn-id: https://hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/flowCore@31424 bc3139a8-67e5-0310-9ffc-ced21a209358

\section{Introduction}

@@

\textit{How does the flow data world look like? Other flow software
  (what other software are handling multiple FCS files at
  all). Somebody more familiar with flowJo and friends should add to
  this...}

{\bf NLM:} Only a couple of commercial software (e.g., FlowJo and
DeNovo) allows the user to analyse batch of FCS files.  Only a few
allows export in XML (e.g., FlowJo).

{\bf RB:}  It would be important to highlight
the differences between flowCore and commercial software packages,
including in philosophy (open vs. closed source) and interaction
(programmatic data analysis vs. GUI-driven).

\section{Basic data structures}
\subsection*{flowFrame}
flowCore's primary task is the representation and basic manipulation
of flow cytometry data. This is accomplished through a data model very
close to that adopted by other Bioconductor packages used for gene
expression analysis, and thus familiar to most Bioconductor
users. Similar to any other flow tool, the basic unit of organization
in flowCore is a collection of events. We call the structure that hold
this data a flowFrame. All relevant information for the collection are
stored along with the matrix of raw data values and can be accessed
programmatically. Most prominently, these are descriptors of the
stains used in the experiment and the respective measurement channels,
information about compensation performed at the instrument side and
any additional keywords the user deems to be important to annotate the
data. A number of quality checks are performed during the creation of
a flowFrame to ensure data integrity. There is an extended interface
that allows for the standardized interaction with the flow data, thus
enabling the development of more specialized tools for specific tasks
of the data analysis process.

@@

\section{Standard flow operations}
The basic operations in flow cytometry analysis are almost always the
same: the data need to be compensated (if not already done on the
instrument side), often transformed and sup-population of interest
need to be selected based on a set of (predominantly sequential)
gates. All software solutions for flow data analysis offer more or
less elaborate support for these kinds of operation, most of them in
an interactive, GUI-driven spirit. The approach we have taken in
flowCore is to abstractly describe these operations and build a set of
tools to perform them on both flowFrames as well as on entire
flowSets. While transformation and to a certain extent also
compensation are fairly routine operations with only limited potential
for improvement, being able to implement new methodology for gating of
flow cytometry data and extending the capabilities of flowCore through
the usual ways of object oriented programming is a feature, that
clearly sets our framework apart from all the other flow toolsets
around. By factoring out as much of the bookkeeping as possible,
programmers can focus on the actual operations rather then having to
deal with the tedious details of data integration and access. Third
party methods implemented by others can act on their own as
first-class citizens in the analysis framework, without breaking the
workflow or the basic infrastructure. This design allows for the
straight forward extension of flowCore's capabilities and already
fostered the development of a number of valuable
improvements. \citep{(flowClust, plateCore)}

@@

\section{Related flow packages}
In addition to the flowCore package that offers basic infrastructure,
we have implemented a range of additional Bioconductor packages that
are dedicated to more specific tasks of flow cytometry data
analysis. The flowViz package provides sophisticated data
visualization tools, that make use of multivariate trellis
plotting. These functions can be used to quickly generate customized
plots for extended cytometry data sets for both direct data inspection
and quality control. The flowQ package offers more advanced quality
assurance (QA) methodology and a framework to create interactive
web-based reports of QA results. Most of flowUtil's content deals with
data import and output from and to flowCore and the flow-cytometry
specific standard markup languages. Finally, flowStats provides
statistical methods that are relevant in the context of flow cytometry
data analysis.

@@

{\bf NLM:} example: cite flowClust and floViz paper?

\section{Application, Outlook}
\textit{This is the most important part, we need to show nice examples
  of use cases, limit ourselfs to the basics and show what we can do
  that nobody else can do at the moment. Need at lot of input here
  from everybody. Also for the things that go in here we should
  provide further details in the Supplementary Materials.}

{\bf NLM:} We could illustrate/summarize:
\begin{itemize}
\item the power of subsetting: for example to manipulate 
patient below a certain age or of the same sex...
\item Gating: filterSet strategy, kmeanFilter
\item visualization adding metadata as parameter
example: cite flowViz paper
\end{itemize}

>>>>

\section*{Introduction}

@@

\section*{Basic data structures}
\subsection*{flowFrame}
flowCore's primary task is the representation and basic manipulation
of flow cytometry data. This is accomplished through a data model very
close to that adopted by other Bioconductor packages used for gene
expression analysis, and thus familiar to most Bioconductor
users. Similar to any other flow tool, the basic unit of organization
in flowCore is a collection of events. We call the structure that hold
this data a flowFrame. All relevant information for the collection are
stored along with the matrix of raw data values and can be accessed
programmatically. Most prominently, these are descriptors of the
stains used in the experiment and the respective measurement channels,
information about compensation performed at the instrument side and
any additional keywords the user deems to be important to annotate the
data. A number of quality checks are performed during the creation of
a flowFrame to ensure data integrity. There is an extended interface
that allows for the standardized interaction with the flow data, thus
enabling the development of more specialized tools for specific tasks
of the data analysis process.

@@

\section*{Standard flow operations}
The basic operations in flow cytometry analysis are almost always the
same: the data need to be compensated (if not already done on the
instrument side), often transformed and sup-population of interest
need to be selected based on a set of (predominantly sequential)
gates. All software solutions for flow data analysis offer more or
less elaborate support for these kinds of operation, most of them in
an interactive, GUI-driven spirit. The approach we have taken in
flowCore is to abstractly describe these operations and build a set of
tools to perform them on both flowFrames as well as on entire
flowSets. While transformation and to a certain extent also
compensation are fairly routine operations with only limited potential
for improvement, being able to implement new methodology for gating of
flow cytometry data and extending the capabilities of flowCore through
the usual ways of object oriented programming is a feature, that
clearly sets our framework apart from all the other flow toolsets
around. By factoring out as much of the bookkeeping as possible,
programmers can focus on the actual operations rather then having to
deal with the tedious details of data integration and access. Third
party methods implemented by others can act on their own as
first-class citizens in the analysis framework, without breaking the
workflow or the basic infrastructure. This design allows for the
straight forward extension of flowCore's capabilities and already
fostered the development of a number of valuable
improvements. \citep{lo2008agf} (+plateCore).

@@

\section*{Related flow packages}
In addition to the flowCore package that offers basic infrastructure,
we have implemented a range of additional Bioconductor packages that
are dedicated to more specific tasks of flow cytometry data
analysis. The flowViz package provides sophisticated data
visualization tools, that make use of multivariate trellis
plotting. These functions can be used to quickly generate customized
plots for extended cytometry data sets for both direct data inspection
and quality control. The flowQ package offers more advanced quality
assurance (QA) methodology and a framework to create interactive
web-based reports of QA results. Most of flowUtil's content deals with
data import and output from and to flowCore and the flow-cytometry
specific standard markup languages. Finally, flowStats provides
statistical methods that are relevant in the context of flow cytometry
data analysis.

@@

\section*{Application}
The flowCore package has already been applied in the analysis of a
number of datasets \citep{gasparetto2004ice,brinkman2007hcf}, and
several other Bioconductor packages are making use of its code
base. As an example, \cite{lo2008agf} have recently proposed an
automatic gating approach via robust model-based clustering using the
flowCore data model and infrastructure (implemented in the
Bioconductor package flowClust). Another package providing more
specialized support for experiments conducted on microtitre plates is
currently being developed. The documentation of all of flowCore's
features is way beyond the scope of this publication. Instead, we
want to use a single example to demonstrate some of the software's
hallmarks, that is, the concept of data-driven automated gating, the
integration of existing software into the framework and the generation
of publication-quality graphics for data visualization. The data used
for this example are flow measurements of human T-lymphocytes from 30
different patients, stained for both the CD4 and CD8 markers.

A very basic matrix of scatterplots is shown in Figure \ref{xyplot},
where each panel in the matrix represents the data of one individual
patient. Instead of plotting individual points, the local bivariate
density of the data is shown as a false color image. For each panel,
the outlines of the regions identified in the automated gating
operation described below are added to the plot. The input to the
plotting function is either a single flow frame or a complete flow set
object, and all available metadata can be accessed to control the
appearance and the composition of the plot. Similarily, gate objects
are directly passed on to the plotting function and are included in
the plot if possible. Besides the very basic scatterplot shown here,
there is a wealth of additional data visualizations available in the
flowViz package\cite{sarkar2008ufv}, all making use of flowCore's data
models allowing to easily arrange the data according to numerous
experimental factors. Furthermore, the design and the API of the
visualization software itself is very generic, and the user can
readily extend its capabilities by providing self-defined plotting
functions.

Static gating for all samples in a high-throughput flow cytometry
experiment is often not an option, since the measured variables tend
to vary between different treatments or over time. Automated or
data-driven gating tries to estimate the gating regions from the
underlying data, thus providing a fast objective solution to the
analysis of potentially very large and diverse data sets
\cite{lo2008agf}. One of the automated gating methods implemented in
flowCore is based on identifying areas of significant curvature in a
kernel density estimate of the data (\ref{Wandetal}). Assuming that
the regions of interest are of high density, the software is able to
reliably detect them in either the one ore the two-dimensional density
landscape. Kernel density estimation is a well-known problem in
statistical computing, and a lot of effort has been invested in the
development of good software to address it(\ref{Wandetal}). The
modular design of R, Bioconductor and flowCore allows to easily
integrate these existing solutions into our framework. Instead of
re-writing existing code, we are able to include it via the well
tested distribution mechanism provided by R's software package
system. This process is bi-directional, and all funtionality
implemented in flowCore is available to other package authors, as we
have seen with the afore-mentioned flowClust package.

In developing flowCore, we tried to provided a unified develoment and
analysis platform for flow-cytometry data. The scope and complexity of
the technology is evolving rapidly, and a collaborative effort in
devising new methodolgy has been proven beneficial for a number of
different biological and computational biology challenges. We hope
that our framework will be the foundation for fruitful shared research
by many calloborators from multiple scientific fields, further pushing
the boundaries of the exiting and still promising flow
cytometry technology.
